---
title: "ESM 244 Lab 1"
author: "Mia Guarnieri"
date: "1/15/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#echo = true means to show the code chunks

library(tidyverse)
library(here)
library(sf)
library(tmap)

#update.packages will update the packages, ask = FALSE means it won't ask for each one
```

### read in the data

cmd - alt - i to insert new code chunk

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### part 1: wrangling and ggplot review

**Example 1:** find counts of observation by `legal_status` & wrangle a bit

```{r}
#method e: group_by() %>%  summarize
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

#method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(tree_count)
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = "darkgreen") +
  labs(x = "Legal Status", y = "Tree Count") +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA" and store as `permitted_data_df`

```{r}
permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", #vertical line instead of a comma means OR
         caretaker == "MTA")
```

**Example 3:** Only keep blackwood acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude`, and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

#make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = "darkgreen")

```

**Example 4:** Use `tidyr :: separate()` 

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::")
```

**Example 5:** Use `tidyr :: unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite("id_status", tree_id, legal_status, sep = "_cool_")
```

### Part 2: make some maps

**Step 1:** Convert the lat/lon to spatial point, `sf::st_as_sf()`

```{r}
blackwood_acacia_sf <- blackwood_acacia_df %>% 
  drop_na(lat, lon) %>% 
  st_as_sf(coords = c("lon", "lat"))

#we need to tell R what the coordinate reference system is
st_crs(blackwood_acacia_sf) <- 4326

ggplot(data = blackwood_acacia_sf) +
  geom_sf(color = "darkgreen") +
  theme_minimal()

```

Read in the SF shapefile and add to map
```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

sf_map_transform <- st_transform(sf_map, 4326)

ggplot(data = sf_map_transform)+
  geom_sf()
```

Combine the maps!

```{r}
ggplot()+ 
  geom_sf(data = sf_map_transform,
          size = 0.1,
          color = "darkgrey")+
  geom_sf(data = blackwood_acacia_sf,
          color = "darkgreen",
          size = 0.5) +
  theme_void() +
  labs(title = "Blackwood acacias in SF")
```

### Now an interactive map

```{r}
tmap_mode("view")

tm_shape(blackwood_acacia_sf) +
  tm_dots()
```




